---
title: "comparative_analysis"
author: "Owen Liu"
date: "June 13, 2016"
output: html_document
---

## Effect of MPAs and Sea Otters on Kelp Forest Benthic Communities

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
library(car)
library(boot)
library(vegan)

# a couple of convenience functions
lunique <- function(x) length(unique(x)) # number of unique values in a variable
datclass <- function(x) lapply(x,class) # variable classes in a dataset

WD <- getwd() # working directory
```

Dataset import from separate data preparation file.  See that file for details on the three large ecological datasets we've brought together.

```{r data import}
dat <- read.csv(file=paste0(WD,"/data/data_complete_6_13_16.csv"),stringsAsFactors = F)

```

As a quick recap, these data are from 3 sources:

* The Channel Islands Kelp Forest Monitoring Program (KFMP)
* The San Nicolas Island/USGS Monitoring Program (SNI)
* The Partnership for the Interdisciplinary Study of the Coastal Ocean (PISCO)

The data contain information on the mean, standard error, and variance in density of 14 species of algae and invertebrates across 32 years and 259 sites, located on the Central and Southern California coast and the California Channel Islands.  Also included in the data are the level and duration of spatial protection afforded to the site (i.e., if there is an MPA).  Finally, the data have information on when California sea otters established residency at particular sites, based on a separate sea otter census dataset.  Given these data, we can explore the effects of both protection and an important top predator on kelp forest benthic community structure and dynamics.

```{r exploratory plots}

# add a CV variable (ratio of variance to mean)
dat <- mutate(dat,CV=sqrt(var.dens)/mean.dens)

# how does mean relate to variance? log scale
ggplot(dat,aes(x=log(mean.dens),y=log(var.dens),col=ID_all))+geom_point()+geom_smooth(method="lm")
summary(lm(log(var.dens)~log(mean.dens),data=filter(dat,var.dens>0)))

# MPA protection
dat.mpa <- filter(dat,!is.na(yrs.prot)) %>% mutate(bef.aft=ifelse(yrs.prot<=0,"Before","After"))
ggplot(filter(dat.mpa,ID_all=="STRP"),aes(x=bef.aft,y=CV)) + geom_boxplot()

# test for difference in CV before/after, purple urchin
dat.mpa.strp <- filter(dat.mpa,ID_all=="STRP")
t.test(dat.mpa.strp$CV[dat.mpa.strp$bef.aft=="Before"],dat.mpa.strp$CV[dat.mpa.strp$bef.aft=="After"])

# test for difference in CV before/after, macrocystis
dat.mpa.macp <- filter(dat.mpa,ID_all=="MACP")
t.test(dat.mpa.macp$CV[dat.mpa.macp$bef.aft=="Before"],dat.mpa.macp$CV[dat.mpa.macp$bef.aft=="After"])

# Otter presence
dat.otter <- filter(dat,!is.na(otteryear)) %>% 
  mutate(yrs.ott=year-otteryear) %>%
  mutate(bef.aft=ifelse(yrs.ott<=0,"Before","After"))
ggplot(filter(dat.otter,ID_all=="STRP"),aes(x=bef.aft,y=CV,fill=island)) + geom_boxplot()
ggplot(filter(dat.otter,ID_all=="MACP"),aes(x=bef.aft,y=CV,fill=island)) + geom_boxplot()

# test for difference in CV before/after
dat.ott.strp <- filter(dat.otter,ID_all=="STRP") %>% filter(CV>0) %>% mutate(logCV=log(CV))
t.test(dat.ott.strp$logCV[dat.ott.strp$bef.aft=="Before"],dat.ott.strp$logCV[dat.ott.strp$bef.aft=="After"])
```

If we want to compare variability across space and time, one thing we have to check is whether mean and variance are related for the species in our study.  We can do this by estimating Taylor power plot slopes for each species (McArdle et al., 1990), or, equivalently, testing the relationship of log(mean) and log(CV).  If $\beta$ is not significantly different than 2, or the slope of the log(mean) to log(CV) plot is not significantly different than 0, then CV is an appropriate measure of variability which is independent of the mean.

```{r slopes and CIs of Taylor plots}
# data where variance is greater than 0
dat2 <- filter(dat,var.dens>0)

# amount of data for each species
specounts <- dat2 %>% group_by(ID_all) %>% count(ID_all)

taylor.coeffs <- data.frame(species=unique(dat2$ID_all),slope=NA,lower=NA,upper=NA,p=NA)
for(i in taylor.coeffs$species) {
  testdat <- filter(dat2,ID_all==i)
  fit <- lm(log(CV)~log(mean.dens),data=testdat)
  slope <- fit$coefficients[2]
  CI <- confint(fit)[2,]
  p <- summary(fit)$coefficients[,4][2]
  taylor.coeffs[taylor.coeffs$species==i,2:5] <- c(slope,CI,p)
}
taylor.coeffs <- taylor.coeffs %>% mutate(sig=ifelse(p<0.05,"Y","N"))

# Basically, these slopes:
ggplot(dat,aes(x=log(mean.dens),y=log(CV),col=ID_all))+
  geom_point()+
  geom_smooth(method="lm")

# Otter binary
dat3 <- dat2 %>% mutate(ott.bin=ifelse(year-otteryear>0,"Yes","No"),bef.aft=ifelse(yrs.prot<=0,"Before","After"))
ggplot(filter(dat3,ID_all=="STRP"),aes(x=bef.aft,y=log(CV),col=ott.bin))+
  geom_boxplot()

# *****************************************
# Variability measured across space (aggregated across time)
dat.temporal <- dat %>% group_by(ID_all,sitelong) %>% 
  summarise(mean=mean(mean.dens),CV=sd(mean.dens)/mean(mean.dens)) %>%
  filter(CV>0)

ggplot(dat.temporal,aes(x=log(mean),y=log(CV),col=ID_all)) + geom_point() + geom_smooth(method="lm")

## Linear coefficients of log(mean)/log(CV) relationship, with 95% bootstrapped CIs and p-value
# Need a function to perform on each bootstrap sample.
bootfun <- function(formula,data,indices) {
  d <- data[indices,] # allows boot to select sample
  fit <- lm(formula,data=d)
  fit$coefficients[2]
}

temporal.coeffs <- data.frame(species=unique(dat.temporal$ID_all),slope=NA,lower=NA,upper=NA,p=NA)
for(i in temporal.coeffs$species) {
  testdat <- filter(dat.temporal,ID_all==i)
  testdat.boot <- boot(data=testdat,statistic=bootfun,R=1000,formula=log(CV)~log(mean))
  slope <- testdat.boot$t0
  CIs <- boot.ci(testdat.boot,type="basic")$basic[4:5]
  fit <- lm(log(CV)~log(mean),data=testdat)
  p <- summary(fit)$coefficients[,4][2]
  temporal.coeffs[temporal.coeffs$species==i,2:5] <- c(slope,CIs,p)
}
temporal.coeffs <- temporal.coeffs %>% mutate(sig=ifelse(p<0.05,"Y","N"))


# Plot of slope results
# change ordering of levels for plotting
temporal.coeffs$species <- factor(temporal.coeffs$species, levels = temporal.coeffs$species[order(desc(temporal.coeffs$slope))])

ggplot(temporal.coeffs,aes(x=species,y=slope)) +
  geom_pointrange(aes(ymin=lower,ymax=upper)) +
  geom_hline(yintercept=0,linetype=2) +
  ggtitle("Slope of Density-CV log-log Relationship: Temporal") +
  xlab("Species") + ylab("Slope of ln(Density)/ln(CV) relationship")

# ******************************************
# Variability measured across time (aggregated across space)
dat.spatial <- dat %>% group_by(ID_all,year) %>%
  summarise(mean=mean(mean.dens),CV=sd(mean.dens)/mean(mean.dens)) %>%
  filter(CV>0)

ggplot(dat.spatial,aes(x=log(mean),y=log(CV))) + 
  geom_point() + 
  geom_smooth(method="lm")


spatial.coeffs <- data.frame(species=unique(dat.spatial$ID_all),slope.s=NA,lower.s=NA,upper.s=NA,p.s=NA)
for(i in spatial.coeffs$species) {
  testdat <- filter(dat.spatial,ID_all==i)
  testdat.boot <- boot(data=testdat,statistic=bootfun,R=1000,formula=log(CV)~log(mean))
  slope <- testdat.boot$t0
  CIs <- boot.ci(testdat.boot,type="basic")$basic[4:5]
  fit <- lm(log(CV)~log(mean),data=testdat)
  p <- summary(fit)$coefficients[,4][2]
  spatial.coeffs[spatial.coeffs$species==i,2:5] <- c(slope,CIs,p)
}
spatial.coeffs <- spatial.coeffs %>% mutate(sig.s=ifelse(p.s<0.05,"Y","N"))

# Plot of slope results
# change ordering of levels for plotting
spatial.coeffs$species <- factor(spatial.coeffs$species, levels = spatial.coeffs$species[order(desc(spatial.coeffs$slope))])

ggplot(spatial.coeffs,aes(x=species,y=slope.s)) +
  geom_pointrange(aes(ymin=lower.s,ymax=upper.s)) +
  geom_hline(yintercept=0,linetype=2) +
  ggtitle("Slope of Density-CV log-log Relationship: Spatial") +
  xlab("Species") + ylab("Slope of ln(Density)/ln(CV) relationship")

## putting both on the same axes
coeffs.both <- bind_cols(temporal.coeffs,select(spatial.coeffs,-species))

ggplot(coeffs.both,aes(x=slope,y=slope.s,col=species))+
  geom_pointrange(aes(ymin=lower.s,ymax=upper.s))+
  geom_errorbarh(aes(xmin=lower,xmax=upper))+
  ggtitle("Spatial and Temporal Mean-CV Relationship")+
  xlab(expression(paste(beta," ","Temporal")))+
  ylab(expression(paste(beta," ","Spatial"))) +
  scale_color_discrete(name="Species") +
  geom_hline(yintercept=0,linetype=2) +
  geom_vline(xintercept=0,linetype=2) +
  geom_text(aes(label="Hotspots",x=-0.3,y=0.4))+
  geom_text(aes(label="Refugia",x=-0.3,y=-0.5))+
  geom_text(aes(label="Productive Sinks",x=0.2,y=-0.5))+
  geom_text(aes(label="Outbreaks",x=0.2,y=0.4))
```

PCA? Within sites across years; Within years across sites

```{r pca}
siteyears <- dat3 %>% group_by(sitelong,ID_all) %>% summarise(nyears=n()) %>% filter(nyears>5)
pcadat <- filter(dat3,sitelong %in% siteyears$sitelong)

pcamat <- acast(pcadat,sitelong~ID_all,mean,na.rm=T,value.var="mean.dens")
# for now, replace NaN with 0...
pcamat[is.nan(pcamat)] <- 0

pcamat.hel <- decostand(pcamat,method="hel")

all.pca <- rda(pcamat.hel)

plot(all.pca,scaling=1,display=c("wa","sp"))

# extract info from the PCA for customized plotting
pca.plotdat <- data.frame(sitelong=row.names(pcamat.hel))

# plotting data
pca.plotdat <- pca.plotdat %>% 
  mutate(PC1=scores(all.pca,scaling=1,choices=1,display="sites")[,1],PC2=scores(all.pca,scaling=1,choices=2,display="sites")[,1]) %>%
  left_join(select(pcadat,sitelong,island)) %>%
  distinct(sitelong) %>%
  mutate(mainland=ifelse(island=="Mainland",1,0))

## the plot
ggplot(pca.plotdat,aes(x=PC1,y=PC2,col=island,shape=factor(mainland)))+
  geom_point(size=3.5) + 
  geom_hline(yintercept=0,linetype=2) +
  geom_vline(xintercept=0,linetype=2) +
  ggtitle("PCA, Benthic Monitoring Sites") +
  guides(shape="none") +
  scale_color_discrete(name="Island")
# 
# ### Thought experiment- bear with me
# spe.pc1 <- scores(all.pca,scaling=1,display="species")[,1]
# sni_3 <- dat3 %>% filter(sitelong=="SNI_3") %>% dcast(year~ID_all,value.var="mean.dens")
# sni_3[is.na(sni_3)] <-0
# sni_3.hel <- sni_3 %>% select(-year) %>% decostand(method="hel")
# sni_3_pc1 <- apply(sni_3.hel,1,function(x)sum(x*spe.pc1))
# sni_3_pc1 <- data_frame(year=sni_3$year,pc1=sni_3_pc1)

```

Treatments - where are there otters and mpas?

```{r treatments}
treat.ott <- filter(dat3,!is.na(otteryear)) %>% group_by(sitelong,ID_all) %>% summarise(n())
treat.mpa <- filter(dat3,!is.na(mpayear)) %>% group_by(sitelong,ID_all) %>% summarise(n())
treat.both <- filter(dat3,!is.na(otteryear)&!is.na(mpayear)) %>% group_by(sitelong,ID_all) %>% summarise(n())

dat4<-dat3
dat4$ott.bin[is.na(dat4$ott.bin)]<-"No"
dat4$yrs.prot[is.na(dat4$yrs.prot)]<-0
summary(lm(CV~mean.dens+island+yrs.prot+ott.bin+dataset+lat,data=dat4))
```