---
title: "comparative_analysis"
author: "Owen Liu"
date: "June 13, 2016"
output: html_document
---

## Effect of MPAs and Sea Otters on Kelp Forest Benthic Communities

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
library(car)
library(boot)
library(vegan)

# a couple of convenience functions
lunique <- function(x) length(unique(x)) # number of unique values in a variable
datclass <- function(x) lapply(x,class) # variable classes in a dataset

WD <- setwd("C:/Users/oliu.MASONBEE/Documents/github/kelp_forest_variability") # working directory
```

Dataset import from separate data preparation file.  See that file for details on the three large ecological datasets we've brought together.

```{r data import}
dat <- read.csv(file=paste0(WD,"/data/data_complete_6_13_16.csv"),stringsAsFactors = F)

```

As a quick recap, these data are from 3 sources:

* The Channel Islands Kelp Forest Monitoring Program (KFMP)
* The San Nicolas Island/USGS Monitoring Program (SNI)
* The Partnership for the Interdisciplinary Study of the Coastal Ocean (PISCO)

The data contain information on the mean, standard error, and variance in density of 14 species of algae and invertebrates across 32 years and 259 sites, located on the Central and Southern California coast and the California Channel Islands.  Also included in the data are the level and duration of spatial protection afforded to the site (i.e., if there is an MPA).  Finally, the data have information on when California sea otters established residency at particular sites, based on a separate sea otter census dataset.  Given these data, we can explore the effects of both protection and an important top predator on kelp forest benthic community structure and dynamics.

```{r exploratory plots}

# add a CV variable (ratio of variance to mean)
dat <- mutate(dat,CV=sqrt(var.dens)/mean.dens)

# how does mean relate to CV? log scale
ggplot(dat,aes(x=log(mean.dens),y=log(CV),col=ID_all))+geom_point()+geom_smooth(method="lm")
summary(lm(log(CV)~log(mean.dens),data=filter(dat,var.dens>0)))

# MPA protection
dat.mpa <- filter(dat,!is.na(yrs.prot)) %>% mutate(bef.aft=ifelse(yrs.prot<=0,"Before","After"))
ggplot(filter(dat.mpa,ID_all=="STRP"),aes(x=bef.aft,y=CV)) + geom_boxplot()

# test for difference in CV before/after, purple urchin
dat.mpa.strp <- filter(dat.mpa,ID_all=="STRP")
t.test(dat.mpa.strp$CV[dat.mpa.strp$bef.aft=="Before"],dat.mpa.strp$CV[dat.mpa.strp$bef.aft=="After"])

# test for difference in CV before/after, macrocystis
dat.mpa.macp <- filter(dat.mpa,ID_all=="MACP")
t.test(dat.mpa.macp$CV[dat.mpa.macp$bef.aft=="Before"],dat.mpa.macp$CV[dat.mpa.macp$bef.aft=="After"])

# Otter presence
dat.otter <- filter(dat,!is.na(otteryear)) %>% 
  mutate(yrs.ott=year-otteryear) %>%
  mutate(bef.aft=ifelse(yrs.ott<=0,"Before","After"))
ggplot(filter(dat.otter,ID_all=="STRP"),aes(x=bef.aft,y=CV,fill=island)) + geom_boxplot()
ggplot(filter(dat.otter,ID_all=="MACP"),aes(x=bef.aft,y=CV,fill=island)) + geom_boxplot()

# Otter binary
dat3 <- dat2 %>% mutate(ott.bin=ifelse(year-otteryear>0,"Yes","No"),bef.aft=ifelse(yrs.prot<=0,"Before","After"))
ggplot(filter(dat3,ID_all=="STRP"),aes(x=bef.aft,y=log(CV),col=ott.bin))+
  geom_boxplot()

# test for difference in CV before/after
dat.ott.strp <- filter(dat.otter,ID_all=="STRP") %>% filter(CV>0) %>% mutate(logCV=log(CV))
t.test(dat.ott.strp$logCV[dat.ott.strp$bef.aft=="Before"],dat.ott.strp$logCV[dat.ott.strp$bef.aft=="After"])
```

If we want to compare variability across space and time, one thing we have to check is whether mean and variance are related for the species in our study.  We can do this by estimating Taylor power plot slopes for each species (McArdle et al., 1990), or, equivalently, testing the relationship of log(mean) and log(CV).  If $\beta$ is not significantly different than 2, or the slope of the log(mean) to log(CV) plot is not significantly different than 0, then CV is an appropriate measure of variability which is independent of the mean.

```{r slopes and CIs of Taylor plots}
# data where variance is greater than 0
dat2 <- filter(dat,var.dens>0)

# amount of data for each species
specounts <- dat2 %>% group_by(ID_all) %>% count(ID_all)

taylor.coeffs <- data.frame(species=unique(dat2$ID_all),slope.s2=NA,lower.s2=NA,upper.s2=NA,p.s2=NA)
for(i in taylor.coeffs$species) {
  testdat <- filter(dat2,ID_all==i)
  fit <- lm(log(CV)~log(mean.dens),data=testdat)
  slope <- fit$coefficients[2]
  CI <- confint(fit)[2,]
  p <- summary(fit)$coefficients[,4][2]
  taylor.coeffs[taylor.coeffs$species==i,2:5] <- c(slope,CI,p)
}
taylor.coeffs <- taylor.coeffs %>% mutate(sig.s2=ifelse(p<0.05,"Y","N"))

# Basically, these slopes:
ggplot(dat,aes(x=log(mean.dens),y=log(CV),col=ID_all))+
  geom_point()+
  geom_smooth(method="lm")

# factor for ordered plotting
taylor.coeffs$species <- factor(taylor.coeffs$species, levels = taylor.coeffs$species[order(desc(taylor.coeffs$slope.s2))])

taylor.all <- ggplot(taylor.coeffs,aes(x=species,y=slope.s2)) +
  geom_pointrange(aes(ymin=lower.s2,ymax=upper.s2)) +
  geom_hline(yintercept=0,linetype=2) +
  ggtitle("Slope of Density-CV log-log Relationship: Within Sites") +
  xlab("Species") + ylab("Slope of ln(Density)/ln(CV) relationship")
taylor.all
ggsave("taylor_all.png",width=8,height=5)


# *****************************************
# Variability measured across space (aggregated across time)
dat.temporal <- dat %>% group_by(ID_all,sitelong) %>% 
  summarise(mean=mean(mean.dens),CV=sd(mean.dens)/mean(mean.dens)) %>%
  filter(CV>0)

ggplot(dat.temporal,aes(x=log(mean),y=log(CV),col=ID_all)) + geom_point() + geom_smooth(method="lm")

## Linear coefficients of log(mean)/log(CV) relationship, with 95% bootstrapped CIs and p-value
# Need a function to perform on each bootstrap sample.
bootfun <- function(formula,data,indices) {
  d <- data[indices,] # allows boot to select sample
  fit <- lm(formula,data=d)
  fit$coefficients[2]
}

temporal.coeffs <- data.frame(species=unique(dat.temporal$ID_all),slope=NA,lower=NA,upper=NA,p=NA)
for(i in temporal.coeffs$species) {
  testdat <- filter(dat.temporal,ID_all==i)
  testdat.boot <- boot(data=testdat,statistic=bootfun,R=1000,formula=log(CV)~log(mean))
  slope <- testdat.boot$t0
  CIs <- boot.ci(testdat.boot,type="basic")$basic[4:5]
  fit <- lm(log(CV)~log(mean),data=testdat)
  p <- summary(fit)$coefficients[,4][2]
  temporal.coeffs[temporal.coeffs$species==i,2:5] <- c(slope,CIs,p)
}
temporal.coeffs <- temporal.coeffs %>% mutate(sig=ifelse(p<0.05,"Y","N"))


# Plot of slope results
# change ordering of levels for plotting
temporal.coeffs$species <- factor(temporal.coeffs$species, levels = temporal.coeffs$species[order(desc(temporal.coeffs$slope))])

taylor.temporal <- ggplot(temporal.coeffs,aes(x=species,y=slope)) +
  geom_pointrange(aes(ymin=lower,ymax=upper)) +
  geom_hline(yintercept=0,linetype=2) +
  ggtitle("Slope of Density-CV log-log Relationship: Temporal") +
  xlab("Species") + ylab("Slope of ln(Density)/ln(CV) relationship")
ggsave("taylor_temporal.png",width=8,height=5)


# ******************************************
# Variability measured across time (aggregated across space)
dat.spatial <- dat %>% group_by(ID_all,year) %>%
  summarise(mean=mean(mean.dens),CV=sd(mean.dens)/mean(mean.dens)) %>%
  filter(CV>0)

ggplot(dat.spatial,aes(x=log(mean),y=log(CV))) + 
  geom_point() + 
  geom_smooth(method="lm")


spatial.coeffs <- data.frame(species=unique(dat.spatial$ID_all),slope.s=NA,lower.s=NA,upper.s=NA,p.s=NA)
for(i in spatial.coeffs$species) {
  testdat <- filter(dat.spatial,ID_all==i)
  testdat.boot <- boot(data=testdat,statistic=bootfun,R=1000,formula=log(CV)~log(mean))
  slope <- testdat.boot$t0
  CIs <- boot.ci(testdat.boot,type="basic")$basic[4:5]
  fit <- lm(log(CV)~log(mean),data=testdat)
  p <- summary(fit)$coefficients[,4][2]
  spatial.coeffs[spatial.coeffs$species==i,2:5] <- c(slope,CIs,p)
}
spatial.coeffs <- spatial.coeffs %>% mutate(sig.s=ifelse(p.s<0.05,"Y","N"))

# Plot of slope results
# change ordering of levels for plotting
spatial.coeffs$species <- factor(spatial.coeffs$species, levels = spatial.coeffs$species[order(desc(spatial.coeffs$slope))])

taylor.spatial <- ggplot(spatial.coeffs,aes(x=species,y=slope.s)) +
  geom_pointrange(aes(ymin=lower.s,ymax=upper.s)) +
  geom_hline(yintercept=0,linetype=2) +
  ggtitle("Slope of Density-CV log-log Relationship: Spatial (Across Sites)") +
  xlab("Species") + ylab("Slope of ln(Density)/ln(CV) relationship")

ggsave("taylor_spatial.png",width=8,height=5)

## putting both on the same axes
coeffs.both <- bind_cols(temporal.coeffs,select(spatial.coeffs,-species))

spat.temp.plot<- ggplot(coeffs.both,aes(x=slope,y=slope.s,col=species))+
  geom_pointrange(aes(ymin=lower.s,ymax=upper.s))+
  geom_errorbarh(aes(xmin=lower,xmax=upper))+
  ggtitle("Spatial and Temporal Mean-CV Relationship")+
  xlab(expression(paste(beta," ","Temporal")))+
  ylab(expression(paste(beta," ","Spatial"))) +
  scale_color_discrete(name="Species") +
  geom_hline(yintercept=0,linetype=2) +
  geom_vline(xintercept=0,linetype=2) +
  geom_text(aes(label="Hotspots",x=-0.3,y=0.4))+
  geom_text(aes(label="Refugia",x=-0.3,y=-0.5))+
  geom_text(aes(label="Boom-Bust",x=0.2,y=-0.5))+
  geom_text(aes(label="Outbreaks",x=0.2,y=0.4)) +
  geom_label(aes(label=species),alpha=0.5,nudge_x=0.05,nudge_y=-0.04) +
  guides(color="none")
spat.temp.plot

ggsave("taylor_2way.png",width=8,height=5)

# Macrocystis seems unique. Check out its graphs
ggplot(filter(dat.spatial,ID_all=="MACP"),aes(x=log(mean),y=log(CV))) + 
  geom_point() + 
  geom_smooth(method="lm") +
  ggtitle("Spatial Mean-CV Relationship, Macrocystis")

ggplot(filter(dat.temporal,ID_all=="MACP"),aes(x=log(mean),y=log(CV))) + 
  geom_point() + 
  geom_smooth(method="lm") +
  ggtitle("Temporal Mean-CV Relationship, Macrocystis")

## Same axes with temporal, spatial (within site) and spatial (across sites)
coeffs.all <- left_join(coeffs.both,taylor.coeffs) %>% mutate(mean.slope = (slope+slope.s+slope.s2)/3)
coeffs.all$species <- factor(coeffs.all$species, levels=coeffs.all$species[order(desc(coeffs.all$mean.slope))])

ggplot(coeffs.all,aes(x=species)) +
  geom_pointrange(aes(y=slope,ymin=lower,ymax=upper),shape=17,col="darkgreen") +
  geom_pointrange(aes(y=slope.s,ymin=lower.s,ymax=upper.s),shape=15,col="darkblue") +
  geom_pointrange(aes(y=slope.s2,ymin=lower.s2,ymax=upper.s2),shape=16,col="darkred") +
  scale_shape_identity(name="",labels=c("Temporal","Spatial (Across Sites)","Spatial (Within Sites)")) +
  geom_hline(yintercept=0,linetype=2) +
  ggtitle("Slope of Density-CV log-log Relationship") +
  xlab("Species") + ylab("Slope of ln(Density)/ln(CV) relationship")


```

PCA? Within sites across years; Within years across sites

```{r pca}
siteyears <- dat3 %>% group_by(sitelong,ID_all) %>% summarise(nyears=n()) %>% filter(nyears>5)
pcadat <- filter(dat3,sitelong %in% siteyears$sitelong)

pcamat <- acast(pcadat,sitelong~ID_all,mean,na.rm=T,value.var="mean.dens")
# for now, replace NaN with 0...
pcamat[is.nan(pcamat)] <- 0

pcamat.hel <- decostand(pcamat,method="hel")

all.pca <- rda(pcamat.hel)

plot(all.pca,scaling=1,display=c("wa","sp"))

# extract info from the PCA for customized plotting
pca.plotdat <- data.frame(sitelong=row.names(pcamat.hel))

# plotting data
pca.plotdat <- pca.plotdat %>% 
  mutate(PC1=scores(all.pca,scaling=1,choices=1,display="sites")[,1],PC2=scores(all.pca,scaling=1,choices=2,display="sites")[,1]) %>%
  left_join(select(pcadat,sitelong,island)) %>%
  distinct(sitelong) %>%
  mutate(mainland=ifelse(island=="Mainland",1,0))

## the plot
ggplot(pca.plotdat,aes(x=PC1,y=PC2,col=island,shape=factor(mainland)))+
  geom_point(size=3.5) + 
  geom_hline(yintercept=0,linetype=2) +
  geom_vline(xintercept=0,linetype=2) +
  ggtitle("PCA, Benthic Monitoring Sites") +
  guides(shape="none") +
  scale_color_discrete(name="Island")
# 
# ### Thought experiment- bear with me
# spe.pc1 <- scores(all.pca,scaling=1,display="species")[,1]
# sni_3 <- dat3 %>% filter(sitelong=="SNI_3") %>% dcast(year~ID_all,value.var="mean.dens")
# sni_3[is.na(sni_3)] <-0
# sni_3.hel <- sni_3 %>% select(-year) %>% decostand(method="hel")
# sni_3_pc1 <- apply(sni_3.hel,1,function(x)sum(x*spe.pc1))
# sni_3_pc1 <- data_frame(year=sni_3$year,pc1=sni_3_pc1)

```

Treatments - where are there otters and mpas?

```{r treatments}
treat.ott <- filter(dat,!is.na(otteryear)) %>% group_by(sitelong,ID_all) %>% summarise(n()) %>% left_join(select(dat,sitelong,otteryear)) %>% distinct(sitelong,ID_all)

treat.mpa <- filter(dat,!is.na(mpayear)) %>% group_by(sitelong,ID_all) %>% summarise(n())%>% 
  left_join(select(dat,sitelong,mpayear)) %>% distinct(sitelong,ID_all)

treat.both <- filter(dat,!is.na(otteryear)&!is.na(mpayear)) %>% group_by(sitelong,ID_all) %>% summarise(n())%>% left_join(select(dat,sitelong,mpayear,otteryear)) %>% distinct(sitelong,ID_all)

dat4<-dat3
dat4$ott.bin[is.na(dat4$ott.bin)]<-"No"
dat4$yrs.prot[is.na(dat4$yrs.prot)]<-0
```

Now we can look at kelp/urchin biplots across treatments

```{r urchins vs kelp}

# Sum of all brown algae by site/year
kelp.dat <- dat4 %>% filter(ID_all %in% c("MACP","PTER","EISA","LAMI")) %>% 
  group_by(sitelong,year,island,ott.bin,yrs.prot) %>%
  summarise(brown.dens=sum(mean.dens))

# Sum of both urchin species by site/year
urch.dat <- dat4 %>% filter(ID_all %in% c("STRF","STRP")) %>% 
  group_by(sitelong,year,island,ott.bin,yrs.prot) %>%
  summarise(urch.dens=sum(mean.dens))

# Joined urchin and kelp data, keeping only matching records
urch.kelp <- inner_join(kelp.dat,urch.dat) %>% mutate(prot.cat=ifelse(yrs.prot<=5,"Less than 5","More than 5"))

## plot of urchin density vs. kelp density
ggplot(urch.kelp,aes(x=urch.dens,y=brown.dens)) + 
  geom_point(aes(col=island,shape=ott.bin)) +
  ggtitle("Urchin Density vs. Brown Algae Density") +
  xlab("Urchin Density (ind./ sq. m)") +
  ylab("Brown Algae Density (ind./sq. m)")

## Just places with otters
ggplot(filter(urch.kelp,ott.bin=="Yes"),aes(x=urch.dens,y=brown.dens)) + 
  geom_point(aes(col=island)) +
  ggtitle("Urchin Density vs. Brown Algae Density") +
  xlab("Urchin Density (ind./ sq. m)") +
  ylab("Brown Algae Density (ind./sq. m)")

## Places without otters
ggplot(filter(urch.kelp,ott.bin=="No"),aes(x=urch.dens,y=brown.dens)) + 
  geom_point(aes(col=island)) +
  ggtitle("Urchin Density vs. Brown Algae Density") +
  xlab("Urchin Density (ind./ sq. m)") +
  ylab("Brown Algae Density (ind./sq. m)")

## MPA protection
ggplot(urch.kelp,aes(x=urch.dens,y=brown.dens)) + 
  geom_point(aes(col=island,shape=prot.cat)) +
  ggtitle("Urchin Density vs. Brown Algae Density") +
  xlab("Urchin Density (ind./ sq. m)") +
  ylab("Brown Algae Density (ind./sq. m)")

## SD/CV urchin density with/without otters, with/without MPA
sd(urch.kelp$urch.dens[urch.kelp$prot.cat=="More than 5"])/mean(urch.kelp$urch.dens[urch.kelp$prot.cat=="More than 5"])
sd(urch.kelp$urch.dens[urch.kelp$prot.cat=="Less than 5"])/mean(urch.kelp$urch.dens[urch.kelp$prot.cat=="Less than 5"])
sd(urch.kelp$urch.dens[urch.kelp$ott.bin=="Yes"])/mean(urch.kelp$urch.dens[urch.kelp$ott.bin=="Yes"])
sd(urch.kelp$urch.dens[urch.kelp$ott.bin=="No"])/mean(urch.kelp$urch.dens[urch.kelp$ott.bin=="No"])


```