---
title: "data_prep"
author: "Owen Liu"
date: "June 3, 2016"
output: html_document
---

## Data Preparation for variability analysis
The objective of this project is to investigate the variability in kelp forest community structure over time and space.

This script will take the data from monitoring projects at San Nicolas Island and PISCO, and summarize their means and variablity for certain key species over space and time.  For now, we are using benthic density data, which, for all datasets, counts organisms like algaes and invertebrates across transects.

We have to split data down to the swath level, pull out the appropriate species, and then aggregate at the appropriate scales.

We'll start with the data from San Nicolas island.

### San Nic data
```{r setup, echo=F,message=F}
library(dplyr)
library(ggplot2)
library(reshape2)
library(tidyr)
library(car)
library(lubridate)

# a couple of convenience functions
lunique <- function(x) length(unique(x)) # number of unique values in a variable
datclass <- function(x) lapply(x,class) # variable classes in a dataset

WD <- getwd() # working directory
```


```{r san nic data, message=F}
## Import raw San Nic data
# *************************
sn.dat <- read.csv(file=paste0(WD,"/data/san nic/Benthic density raw data.csv"),stringsAsFactors = F)
# Dates and years
sn.dat <- sn.dat %>% mutate(date=as.Date(Date,format="%m/%d/%y"),Date=NULL,year=year(date),month=month(date),day=day(date))

# species name/code key
sn.spp.key <- read.csv(file=paste0(WD,"/data/san nic/Table4_Species_sampled.csv"),stringsAsFactors = F)
# just species sampled in the benthic density swaths
sn.spp.key <- sn.spp.key %>% filter(DataSet=="Benthic density") %>% select(-DataSet) %>% mutate(dat="SNI")

# List to store all the data (for calculating variance later)
sn.spp.list <- list()

# Break data down to the species/year level.
for (i in unique(sn.spp.key$SpeciesCode)) {
  
  datname <- paste0("sn.spp.",i) # name for the output
  
  dat <- sn.dat %>% filter(SpeciesCode==i) %>% # subset the species, order by station and swath, and join species name to it
    arrange(Station,Swath) %>% left_join(sn.spp.key)
  
  assign(datname,dat) # assign data for that species to a variable
  
  sn.spp.list[[datname]] <- dat
}
```

Variables in these data for each species are:  
* Station: Site identifier. There are individual swaths within each station.
* Period: Sampling period.  Each sampling period corresponds to one field excursion, but may encompass several actual dates (i.e., a sampling period may last a week)  
* Date
* Swath: Identified by distance along a 50m main transect, and "L" or "R" indicating whether the transect is perpendicular to the left or right of the main line.
* SpeciesCode and SpeciesName
* Density: Individuals per square meter.


### PISCO Data
PISCO has a different sampling design, and the data recorded are in counts, instead of pre-transformed to densities.  We'll calculate density by dividing individual counts in each record by 60 (because organisms are counted on 30m x 2m swaths)
```{r pisco data,message=F}
# read in dataset
ps.dat <- read.csv(file=paste0(WD,"/data/pisco/pisco_subtidal.161.2.csv"),stringsAsFactors = F)
ps.dat <- ps.dat %>% select(-notes,-observer,-depth,-campus,-method) # removing unneeded columns

# site key
ps.site.key <- read.csv(file=paste0(WD,"/data/pisco/pisco_subtidal.181.1.csv"),stringsAsFactors = F)
ps.site.key <- ps.site.key %>% filter(method=="SBTL_SWATH")# just swath sampling

# species key
ps.spp.key <- read.csv(file=paste0(WD,"/data/pisco/pisco_subtidal.180.1.csv"),stringsAsFactors = F)
ps.spp.key <- ps.spp.key %>% filter(sample_type=="SWATH",!(pisco_classcode=="NO_ORG"|pisco_classcode=="ASTSPP"|pisco_classcode=="CRYDEC"|pisco_classcode=="GORGAD"|pisco_classcode=="UNIDSP")) %>%
  select(-sample_type,-sample_subtype) %>% mutate(dat="PISCO")
# just swath sampling and removing "non-organisms"

# List to store all the data (for calculating variance later)
ps.spp.list <- list()

# Break data down to the species/year level.
for (i in unique(ps.spp.key$pisco_classcode)) {
  
  datname <- paste0("ps.spp.",i) # name for the output
  
  dat <- ps.dat %>% filter(classcode==i) %>% # subset the species, join species name to it
    left_join(ps.spp.key,by=c("classcode"="pisco_classcode"))
  
  assign(datname,dat) # assign data for that species to a variable
  
  ps.spp.list[[datname]] <- dat
}
```


### Channel Islands Data
This dataset is from the Channel Islands Kelp Forest Monitoring Program (KFMP), Kushner et al. (2013).  Again, data from the KFMP is sampled and monitored a little differently from either the San Nic or PISCO data, and monitoring methods have changed over time.  However, similar to data in the San Nic set, densities are reported for each species at each site in each year, and we can pull out species data in a similar manner:

```{r KFMP data,message=F}
# read in dataset
ci.dat <- read.csv(file=paste0(WD,"/data/channel islands/benthdensdata.csv"),stringsAsFactors = F)

# Fix dates
ci.dat <- ci.dat %>% mutate(date=as.Date(Date,format="%d-%b-%Y"),year=year(date),month=month(date),day=day(date),Date=NULL)

# there is no variance for these data for each site/year combo,but we do have the number of samples, the mean, and the standard error, so we "back-calculate" the sample variance as follows:
# var = (sqrt(n)*SE))^2, where n is the number of replicates, SE is the standard error, and therefore (sqrt(n)*SE) is the sample standard deviation.
# WARNING: THIS IS A BIG ASSUMPTION OF EQUALITY OF SAMPLING METHODS ACROSS DATASETS
ci.dat <- ci.dat %>% mutate(DensityVar=(sqrt(Replicates)*DensitySE)^2)

# site key
ci.site.key <- read.csv(file=paste0(WD,"/data/channel islands/sitesdata.csv"),stringsAsFactors = F)
ci.site.key <- ci.site.key %>% select(-Video,-StartYear,-LoggerDepth__m_) # removing unneeded columns

# species key
ci.spp.key <- read.csv(file=paste0(WD,"/data/channel islands/benthsppid.csv"),stringsAsFactors = F) %>% mutate(dat="KFMP")

# List to store all the data (for calculating variance later)
ci.spp.list <- list()

# Break data down to the species/year level.
for (i in unique(ci.spp.key$species_id)) {
  
  datname <- paste0("ci.spp.",i) # name for the output
  
  dat <- ci.dat %>% filter(Species==i) %>% # subset the species, join species name to it
    left_join(ci.spp.key,by=c("Species"="species_id"))
  
  assign(datname,dat) # assign data for that species to a variable
  
  ci.spp.list[[datname]] <- dat
}
```

### Determining overlapping species
In order to compare these data at all, we need to figure out which species are overlapping
```{r species monitored comparison}
# mush together pisco genus and species to compare to other data
ps.spp.key <- ps.spp.key %>% mutate(gen_spe = paste(genus,species))

# only keep overlaps
spp.key.all <- ps.spp.key %>% 
  inner_join(ci.spp.key,by=c("gen_spe"="taxonomy")) %>% 
  inner_join(sn.spp.key,by=c("gen_spe"="SpeciesName"))

# unmatched. This way we can investigate name mismatches which may have led to falsely excluded data
spp.key.anti <- ps.spp.key %>% 
  full_join(ci.spp.key,by=c("gen_spe"="taxonomy")) %>% 
  full_join(sn.spp.key,by=c("gen_spe"="SpeciesName")) %>%
  anti_join(spp.key.all) %>% arrange(desc(gen_spe))

# seems that we missed macrocystis and megastrea/lithopoma undosa, because of naming differences
spp.key.all <- spp.key.all %>% bind_rows(filter(spp.key.anti,gen_spe==("Macrocystis pyrifera")|gen_spe==("Macrocystis pyrifera <1m")|gen_spe==("Macrocystis pyrifera >1m")|gen_spe=="Megastrea undosa"|gen_spe=="Megastrea undosum"|gen_spe=="Laminaria spp."|gen_spe=="Laminaria farlowii"))

spp.key.all$gen_spe
```

Here's our final list of species common to all three datasets.

A quick test with one species, to see if data is comparable

```{r red urchin gut check}
# data from pisco
ps.red <- select(ps.spp.STRFRAAD,density) %>%mutate(ds="PISCO",Density=density) %>% select(-density)

# data from KFMP
ci.red <- select(ci.spp.11005,DensityMean) %>% mutate(ds="KFMP",Density=DensityMean) %>% select(-DensityMean)

# data from SNI
sn.red <- select(sn.spp.29,Density) %>% mutate(ds="SNI")

# bind all together for stat tests and plotting
redtest <- bind_rows(ps.red,ci.red,sn.red)

# boxplot of distributions
ggplot(redtest,aes(x=ds,y=Density,col=ds)) + 
  geom_boxplot()+ggtitle("Red Urchin Density Across Datasets") +
  scale_color_discrete(name="Dataset") +xlab("") +
  ylab("Density (ind./square m)")

# Brown-Forsythe test for unequal variance
data.frame(dataset=unique(redtest$ds),variance=c(var(ps.red$Density),var(ci.red$Density,na.rm=T),var(sn.red$Density)),
           n=c(nrow(ps.red),nrow(filter(ci.red,!is.na(Density))),nrow(sn.red)))
leveneTest(Density~factor(ds),data=redtest) # significantly different variance

# Tukey HSD
TukeyHSD(aov(Density~ds,data=redtest))
```



```{r macrocystis gut check}
# data from pisco
ps.mac <- select(ps.spp.MACPYRAD,density) %>%mutate(ds="PISCO",Density=density) %>% select(-density)

# data from KFMP
ci.mac <- select(ci.spp.2002,DensityMean) %>% mutate(ds="KFMP",Density=DensityMean) %>% select(-DensityMean)

# data from SNI
sn.mac <- select(sn.spp.589,Density) %>% mutate(ds="SNI")

# bind all together for stat tests and plotting
mactest <- bind_rows(ps.mac,ci.mac,sn.mac)

# boxplot of distributions
ggplot(mactest,aes(x=ds,y=Density,col=ds)) + 
  geom_boxplot()+ggtitle("Adult Macrocystis Density Across Datasets") +
  scale_color_discrete(name="Dataset") +xlab("") +
  ylab("Density (ind./square m)")

# Brown-Forsythe test for unequal variance
data.frame(dataset=unique(mactest$ds),variance=c(var(ps.mac$Density,na.rm=T),var(ci.mac$Density,na.rm=T),var(sn.mac$Density,na.rm=T)),n=c(nrow(ps.mac),nrow(filter(ci.mac,!is.na(Density))),nrow(sn.mac)))
leveneTest(Density~factor(ds),data=mactest) # significantly different variance

# Tukey HSD
TukeyHSD(aov(Density~ds,data=mactest))
```


### Computing means and variances for each site/year
We need to find a way to pull together the datasets for comparison.  This will require defining a "site" for each dataset and calculating mean and variance in densities for each year at each site.  The process will be a bit different for each dataset, but in order to make them comparable, we'll produce common language:
* dataset: origin of data
* species
* year/month/day
* site - defined differently for each dataset
* n - number of samples contributing to each calculation
* mean.dens - mean density of that species at that site in that year, across swaths/replicates
* se.dens - standard error in density
* var.dens - variance in density

```{r join datasets}


```